Towards a Verified Tableau Prover for Multi-Level-Syllogistics

### 2022-08-30 paper/paper.tex
Continue

@@

definition vars_branch :: 'a branch $\Rightarrow$ 'a set 
  vars_branch b $\equiv$ $\bigcup$(vars_fm ` set b)
\end{lstlisting}
In the standard tableau calculus for propositional logic, a branch is called \textit{closed} if it contains both the negation of a formula and the formula itself;
conversely, it is called \textit{open} if it is not closed.
For MLSS, we extend the notion of closedness with three additional rules; the first two are straightforward while the last one states that a branch is closed when the branch contains a membership cycle \lstinline!s $\inS$ t$_\text{1}$, t$_\text{1}$ $\inS$ t$_\text{2}$, $\ldots$, t$_\text{k}$ $\inS$ s!.
\begin{lstlisting}
inductive bclosed :: 'a branch $\Rightarrow$ bool where
  $\llbracket$ $\phi$ $\in$ set b; &\fmNeg{$\phi$}& $\in$ set b $\rrbracket$ $\Longrightarrow$ bclosed b
| (t $\inS$ $\emptyset$) $\in$ set b $\Longrightarrow$ bclosed b
| (t $\neqS$ t) $\in$ set b $\Longrightarrow$ bclosed b
| $\llbracket$ member_cycle cs; set cs $\subseteq$ set b$\rrbracket$
    $\Longrightarrow$ bclosed b

@@

\section{Completeness of the Calculus}
\begin{itemize}
  \item Basic definitions and graph
  \item Graph is basically a syntactic model.
\end{itemize}
\begin{lstlisting}
definition wf_branch b $\equiv$ $\exists$$\phi$. &\expandss{b}{[$\phi$]}&
\end{lstlisting}

\begin{lstlisting}
definition params :: 'a branch $\Rightarrow$ 'a set where
  params b $\equiv$ vars_branch b - vars_fm (last b)
\end{lstlisting}

\begin{lstlisting}
definition params' :: 'a branch $\Rightarrow$ 'a set where
  params' b $\equiv$
    {c $\in$ params b. $\forall$t $\in$ subterms_fm (last b).
      AT (Var c $\eqS$ t) $\notin$ set b $\land$
      AT (t $\eqS$ Var c) $\notin$ set b} 
\end{lstlisting}

\begin{lstlisting}
definition subterms_branch' ::
    'a branch $\Rightarrow$ 'a pset_term set where
  subterms_branch' b $\equiv$
    subterms_fm (last b) $\cup$ 
    Var ` (params b - params' b)
\end{lstlisting}

\begin{lstlisting}
definition bgraph $\equiv$ $\ldots$ 
\end{lstlisting}

@@

@@

\subsection{Formalising the Key Technical Lemma}
Only one quite trivial invariant needed

>>>>

% Record brackets
\usepackage{stmaryrd}

@@

definition vars_branch :: 'a branch $\Rightarrow$ 'a set 
  vars_branch b $\equiv$ $\bigcup$(vars_fm ` set b)
\end{lstlisting}
In the standard tableau calculus for propositional logic, a branch is called \textit{closed} if it contains both the negation of a formula and the formula itself;
conversely, it is called \textit{open} if it is not closed.
For MLSS, we extend the notion of closedness with three additional rules; the first two are straightforward while the last one states that a branch is closed when the branch contains a membership cycle
\begin{center}
\lstinline!s $\inS$ t$_\text{1}$, t$_\text{1}$ $\inS$ t$_\text{2}$, $\ldots$, t$_\text{k}$ $\inS$ s!.
\end{center}
\begin{lstlisting}
inductive bclosed :: 'a branch $\Rightarrow$ bool where
  $\llbracket$ $\phi$ $\in$ set b; &\fmNeg{$\phi$}& $\in$ set b $\rrbracket$ $\Longrightarrow$ bclosed b
| (t $\inS$ $\emptyset$) $\in$ set b $\Longrightarrow$ bclosed b
| (t $\neqS$ t) $\in$ set b $\Longrightarrow$ bclosed b
| $\llbracket$ member_cycle cs; set cs $\subseteq$ set b$\rrbracket$
    $\Longrightarrow$ bclosed b

@@

\section{Completeness of the Calculus}
Once a branch \lstinline!b! is saturated and open at the same time, we are unable to expand the branch further in our calculus.
Thus, to establish completeness of the calculus, we must construct a model \lstinline!M! such that \lstinline!M $\models$ $\phi$! for all formulae \lstinline!$\phi$ $\in$ set b!.
Our model is defined in terms of the following three collections.
\begin{lstlisting}
definition params :: 'a branch $\Rightarrow$ 'a set where
  params b $\equiv$ vars_branch b - vars_fm (last b)

definition params' :: 'a branch $\Rightarrow$ 'a set where
  params' b $\equiv$
    {c $\in$ params b. $\forall$t $\in$ subterms_fm (last b).
      AT (Var c $\eqS$ t) $\notin$ set b $\land$
      AT (t $\eqS$ Var c) $\notin$ set b}

definition subterms_branch' ::
    'a branch $\Rightarrow$ 'a pset_term set where
  subterms_branch' b $\equiv$
    subterms_fm (last b) $\cup$ 
    Var ` (params b - params' b)
\end{lstlisting}
The goal is to construct a syntactic model that is derived from the membership literals \lstinline!(s $\inS$ t) $\in$ set b!.
To this end, we construct a graph whose vertices are set terms and there is an edge from \lstinline!s! to \lstinline!t! in the graph if, and only if, \lstinline!(s $\inS$ t) $\in$ set b!.
Note that we use \citeauthor{graph_theory_afp}'s graph library~\cite{graph_theory_afp} where a graph is represented as a record of vertices, arcs (directed edges), and two functions \lstinline!tail! and \lstinline!head! that map an arc to its source respectively target vertex.
\begin{lstlisting}
definition bgraph b $\equiv$
  let vs = Var ` params' b $\cup$ subterms_branch' b
  in $\llparenthesis$ verts = vs,
     &$\phantom{\llparenthesis}$& arcs = {(s, t). (s $\inS$ t) $\in$ set b},
     &$\phantom{\llparenthesis}$& tail = fst, head = snd $\rrparenthesis$
\end{lstlisting}
The model is defined in terms of a \textit{realization} function relative to this graph.
Observe that the vertices are partitioned into two sets, a collection of parameters \lstinline!Var ` params' b! and the rest of the set terms \lstinline!subterms_branch' b!, which are treated differently by the realization.
Terms in the latter set are evaluated in accordance to the structure of the graph, i.e.\ the realization of a vertex is defined as the union of the realizations of the parent vertices. 
For the former set, we fix a function \lstinline!I! that assigns the elements \lstinline!x $\in$ Var ` params' b! pairwise distinct sets with cardinality greater than that of the vertices and then return the singleton set \lstinline!vset {I x}!.
The cardinality constraint ensures that realizations are distinct between those two sets.  
\begin{lstlisting}
abbreviation parents G s $\equiv$ {u. u $\rightarrow_\text{G}$ s}

function realize :: 'a pset_term $\Rightarrow$ V where
  x $\in$ Var ` params' b
  $\Longrightarrow$ realize x = vset {I x}
| x $\in$ subterms_branch' b
  $\Longrightarrow$ realize t =
  &$\phantom{\Longrightarrow}$&   vset {realize ` parents (bgraph b) s}
| x $\notin$ verts G $\Longrightarrow$ realize x = 0
\end{lstlisting}
Since the function is recursive, we need to ensure that it terminates for it to be well-defined.
The assumption that \lstinline!b! is open implies that there are no membership cycles and thus \lstinline!bgraph b! is acyclic.
Together with the fact that \lstinline!bgraph b! is finite, we conclude that we can only recurse a finite number of times until we encounter a vertex that has no parents.

\subsection{Formalising the Key Technical Lemma}
\begin{lstlisting}
fun is_literal :: 'a fm $\Rightarrow$ bool where
  is_literal (&\fmAtom& _) = True
| is_literal (&\fmNeg{(\fmAtom{} \_)}&) = True
| is_literal _ = False
\end{lstlisting}
\begin{lstlisting}
lemma
  assumes wf_branch b
  assumes c $\in$ params' b
      and t $\in$ params_subterms b
  shows (Var c $\eqS$ t) $\notin$ set b
    and (t $\eqS$ Var c) $\notin$ set b
    and (t $\inS$ Var c) $\notin$ set b
\end{lstlisting}
Only one quite trivial invariant needed

@@

\begin{lstlisting}
definition wf_branch b $\equiv$ $\exists$$\phi$. &\expandss{b}{[$\phi$]}&
\end{lstlisting}

@@
